Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-05-09 article-models-ttderived.tex
reorganization: move morphisms over morphisms to the later subsection
initiated a investigation of a stronger version of inductive morphism

\begin{lem}\label{comp-es}
The following inference rules are valid:
\begin{align*}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq
      {\Gamma}
      {\subst{\tmext{A}{P}{x}{u}}{Q}}
      {\subst{u}{{x}{Q}}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    }
  { \jtermeq
      {\Gamma}
      {\subst{u}{{x}{Q}}}
      {\subst{\tmext{A}{P}{x}{u}}{g}}
      {\subst{u}{{x}{g}}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jfam{{{{\Gamma}{A}}{P}}{Q}}{R}
    }
  { \jfameq
      {{\Gamma}{\subst{u}{{x}{Q}}}}
      {\subst{\tmext{A}{P}{x}{u}}{R}}
      {\subst{u}{{x}{R}}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jterm{{{{\Gamma}{A}}{P}}{Q}}{R}{t}
    }
  { \jtermeq
      {{\Gamma}{\subst{u}{{x}{Q}}}}
      {\subst{u}{{x}{R}}}
      {\subst{\tmext{A}{P}{x}{u}}{t}}
      {\subst{u}{{x}{t}}}
    }
\end{align*}
\end{lem}

@@

\subsection{Inductive morphisms}
\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ be a context morphism. We say that $f$ is an 
\emph{inductive morphism} if the following inference rules are valid:
\begin{align*}
& \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jterm{{\Gamma}{B}}{Q}{\jtcext{g}}
    }
& & \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jtermeq{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{\jcomp{A}{f}{\jtcext{g}}}{g}
    }
\end{align*}
\end{defn}

@@

\begin{proof}
The proof is a rather long computation:
\begin{align*}
\jcomp{}{e}{\jcomp{}{e}{P}}
& \jdeq
  \subst{e}
    {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { \ctxwk
        {\jcomp{}{e}{P}}{\subst{e}
        {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\jcomp{}{e}{P}}{{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\ctxext{{A}{P}}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \subst
    {\unfold{\jvcomp{\jcomp{}{e}{P}}{e}{\idtm{\jcomp{}{e}{P}}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}
  \\
& \jdeq 
  \jcomp{}{\jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{
    \jcomp{}{\jvcomp{}{\idtm{A}}{f}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{\idtm{A}}{f}}{
    \jcomp{}{e}{P}}
  \\
& \jdeq
  \subst
    {\tmext{\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}{f}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    { {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    { {\cprojfstf{A}{\ctxext{P}{\jcomp{}{e}{P}}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    {\jcomp{}{\cprojfstf{A}{\ctxext{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{f}{\jcomp{}{e}{P}}.
  \qedhere
\end{align*}
\end{proof}

@@

@@

>>>>

\begin{lem}\label{comp-es}
The following inference rules are valid:
\begin{align*}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq
      {\Gamma}
      {\subst{\tmext{A}{P}{x}{u}}{Q}}
      {\subst{u}{{x}{Q}}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jfam{{{{\Gamma}{A}}{P}}{Q}}{R}
    }
  { \jfameq
      {{\Gamma}{\subst{u}{{x}{Q}}}}
      {\subst{\tmext{A}{P}{x}{u}}{R}}
      {\subst{u}{{x}{R}}}
    }
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{\Gamma}{\subst{x}{P}}{u}
    \jterm{{{{\Gamma}{A}}{P}}{Q}}{R}{t}
    }
  { \jtermeq
      {{\Gamma}{\subst{u}{{x}{Q}}}}
      {\subst{u}{{x}{R}}}
      {\subst{\tmext{A}{P}{x}{u}}{h}}
      {\subst{u}{{x}{h}}}
    }
\end{align*}
\end{lem}

@@

\subsection{Inductive morphisms}
In this subsection we investigate the notion of inductive morphisms. An 
inductive morphism $f$ from $A$ to $B$ in context $\Gamma$ is a morphism which
induces an operation which is judgmentally the right inverse of composition
with $f$. We explore a notion of inductive morphism which is much stronger
than the usual notion: it pushes families over $A$ to families over $B$,
families over families over $A$ to families over families over $B$ and terms
thereof to terms of the output families over families in a manner compatible
with the empty family, extension, weakening, substitution and identity terms.
As a result, inductive morphisms will be stable under extension, weakening,
substitution and the identity term is an inductive morphism.

\begin{defn}
Let $\jhom{\Gamma}{A}{B}{f}$ be a context morphism. We say that $f$ is an 
\emph{inductive morphism} if the following inference rules are valid:
\begin{align*}
& \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jterm{{\Gamma}{B}}{Q}{\jtcext{g}}
    }
& & \inference
  { \jfam{{\Gamma}{B}}{Q}
    \jterm{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{g}
    }
  { \jtermeq{{\Gamma}{A}}{\jcomp{A}{f}{Q}}{\jcomp{A}{f}{\jtcext{g}}}{g}
    }
\end{align*}
\end{defn}

@@

\begin{proof}
The proof is a rather long computation. It was found by starting with the
application of \autoref{eq:ealg-eq1} and making the computation in both ways
from there. The judgmental equality below where \autoref{eq:ealg-eq1} is applied,
is marked with an $*$. The computation goes as follows:
\begin{align*}
\jcomp{}{e}{\jcomp{}{e}{P}}
& \jdeq
  \subst{e}
    {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { \ctxwk
        {\jcomp{}{e}{P}}{\subst{e}
        {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\jcomp{}{e}{P}}{{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\ctxext{{A}{P}}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \subst
    {\idtm{\jcomp{}{e}{P}}}
    { {\ctxwk{\jcomp{}{e}{P}}{e}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \subst
    {\unfold{\jvcomp{\jcomp{}{e}{P}}{e}{\idtm{\jcomp{}{e}{P}}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}
  \\
& \jdeq
  \jcomp{}{\jcomp{}{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}{e}}{P}
  \\
& \stackrel{*}{\jdeq}
  \jcomp{}{
    \jcomp{}{\jvcomp{}{\idtm{A}}{f}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{\idtm{A}}{f}}{
    \jcomp{}{e}{P}}
  \\
& \jdeq
  \subst
    {\tmext{\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}{f}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    { {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    { {\cprojfstf{A}{\ctxext{P}{\jcomp{}{e}{P}}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    {\jcomp{}{\cprojfstf{A}{\ctxext{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{f}{\jcomp{}{e}{P}}.
  \qedhere
\end{align*}
\end{proof}

@@

\begin{proof}
The proof is a straightforward calculation (start in the middle):
\begin{align*}
\subst{{y_2}{{y_1}{\jcomp{}{e}{f}}}}{{y_0}{f}}
& \jdeq
  \subst{\tmext{y_0}{\subst{y_2}{{y_1}{\jcomp{}{e}{f}}}}}{f}
  \\
& \jdeq
  \subst{{y_2}{{y_1}{{y_0}{\jvcomp{}{\idtm{P}}{\jcomp{}{e}{f}}}}}}{f}
  \\
& \jdeq
  \subst
    {y_2}
    { {y_1}
      { {y_0}
        {\jcomp{}{\jvcomp{}{\idtm{P}}{\jcomp{}{e}{f}}}{f}}
        }
      }
  \\
& \jdeq
  \subst
    {y_2}
    { {y_1}
      { {y_0}
        {\jcomp{}{\jvcomp{}{f}{\idtm{\jcomp{}{f}{{}{e}{P}}}}}{f}}
        }
      }
  \\
& \jdeq
  \subst{{y_2}{{y_1}{{y_0}{\jvcomp{}{f}{\idtm{\jcomp{}{f}{{}{e}{P}}}}}}}}{f}
  \\
& \jdeq
  \subst{\tmext{\subst{y_1}{{y_0}{f}}}{y_2}}{f}
  \\
& \jdeq
  \subst{y_2}{{{y_1}{{y_0}{f}}}{f}}.\qedhere
\end{align*}
\end{proof}

@@

\begin{defn}
Let $(A_0,P_0,x_0,f_0)$ be an empty algebra in context $\Gamma$ and consider
\begin{enumerate}
\item a family $\jfam{{\Gamma}{A_0}}{A_1}$.
\item a family $\jfam{{{{\Gamma}{A_0}}{P_0}}{\ctxwk{P_0}{A_1}}}{P_1}$.
\end{enumerate}
\end{defn}
